---
title: "Chapter 2 Team Exercise"
author: "Victor and Meg"
date: "27/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load dataset
data(prostate, package="faraway")
# load required packages
library(tidyverse)
library(ggplot2)
```

## Exploratory Data Analysis

Begin with some basic exploratory data analysis (EDA)

### Data dictionary

Faraway prostate dataset 
(source: https://rdrr.io/cran/faraway/man/prostate.html)

* **lcavol** log(cancer volume)
* **lweight** (prostate weight)
* **age** age
* **lbph** log(benign prostatic hyperplasia amount)
* **svi** seminal vesicle invasion [0=no, 1=yes]
* **lcp** log(capsular penetration)
* **gleason** Gleason score
* **pgg45** percentage Gleason scores 4 or 5
* **lpsa** log(prostate specific antigen)
<br>
<br>

Start by examining the dataset
```{r <EDA>, <options>}
dim(prostate) # dimensions of vector
str(prostate) # display structure
head(prostate) # first parts of object
summary(prostate) # summary statistics of variables, including min and max values
```

<br>
Are there any missing values?

```{r <EDA_3}
# examine missingness
table(is.na(prostate))
```  
* There are no missing data in the entire dataset.

<br>

### Graphical outputs

#### Using log(PSA) as outcome value, the following variables appear to be positively correlated:

```{r, results='asis', message = FALSE, error = FALSE, warning = FALSE}
# log(cancer volume)
ggplot(data = prostate) + 
  geom_point(mapping = aes(x = prostate$lcavol, y = prostate$lpsa)) +
  labs(title = 'Plot of log(cancer volume) against log(PSA)', x = 'log(cancer volume)', y = 'log(PSA)')

# log(prostate weight)
ggplot(data = prostate) + 
  geom_point(mapping = aes(x = prostate$lweight, y = prostate$lpsa)) +
  labs(title = 'Plot of log(prostate weight) against log(PSA)', x = 'log(prostate weight)', y = 'log(PSA)')

# gleason score
ggplot(data = prostate) + 
  geom_point(mapping = aes(x = prostate$gleason, y = prostate$lpsa)) +
  labs(title = 'Plot of gleason score against log(PSA)', x = 'gleason score', y = 'log(PSA)')

# log(capsule penetration)
ggplot(data = prostate) + 
  geom_point(mapping = aes(x = prostate$lcp, y = prostate$lpsa)) +
  labs(title = 'Plot of log(capsule penetration) against log(PSA)', x = 'log(capsule penetration)', y = 'log(PSA)')
```

* Note there are a lot of negative log capsular penetration values
* Non negative values do seem to have positive correlation

<br>
<br>

#### Using log(PSA) as outcome value, the following variables appear to have no correlation:

```{r, results='asis', message = FALSE, error = FALSE, warning = FALSE}
# age
ggplot(data = prostate) + 
  geom_point(mapping = aes(x = prostate$age, y = prostate$lpsa)) +
  labs(title = 'Plot of age against log(PSA)', x = 'age', y = 'log(PSA)')

# percentage gleason scores 4 or 5
ggplot(data = prostate) + 
  geom_point(mapping = aes(x = prostate$pgg45, y = prostate$lpsa)) +
  labs(title = 'Plot of % Gleason scores 4 or 5 against log(PSA)', x = '% Gleason scores 4 or 5', y = 'log(PSA)')

# log(BPH)
ggplot(data = prostate) + 
  geom_point(mapping = aes(x = prostate$lbph, y = prostate$lpsa)) +
  labs(title = 'Plot of log(BPH) against log(PSA)', x = 'log(BPH)', y = 'log(PSA)')
```
<br>
<br>

#### Lastly, examining the only dichotomous variable against the outcome (log(PSA))

```{r, results='asis', message = FALSE, error = FALSE, warning = FALSE}
# SVI == 0 (no) vs SVI == 1 (yes) against outcome
ggplot(data=prostate, mapping=aes(x=prostate$lpsa, y=..density..)) + 
  geom_freqpoly(mapping = aes(colour=prostate$svi==0)) +
  labs(title = 'Plot of log(PSA) values with and without presence of seminal vesicle invation', x = 'log(PSA)', y = 'density', colour='Absence of SVI')
```

* No seminal vesicle invasion trended lower log(PSA) than presence of seminal vesicle invasion
<br>

## Fitting linear models
<br>

#### Fit a linear model for outcome log(PSA) and log(cavol) as the single predictor
```{r <linear_models>, <options>}
# fits a linear model, with the `lpsa` as the outcome, and `lcavol` as the single predictor
lcavol_mod <- lm(lpsa ~ lcavol, data=prostate)

# store the R-squared and RSE values
lcavol_mod_summ <- summary(lcavol_mod)
lcavol_rsq <- lcavol_mod_summ$adj.r.squared
lcavol_RSE <- lcavol_mod_summ$sigma

# display lcavol model R-squared and RSE
lcavol_mod_output <- data.frame(predictor="lcavol", R_squared=lcavol_rsq , RSE=lcavol_RSE)
print(lcavol_mod_output)
```
<br>

#### Add each of the variables `lweight`, `svi`, `lbph`, `age`, `lcp`, `pgg45` and `gleason` to the model, one-by-one
```{r <linear_models_2>, <options>}
# fit multiple linear models, adding one more variable successively to each model
lcavol_lweight_mod <- lm(lpsa ~ lcavol + lweight, data=prostate)
lcavol_lweight_svi_mod <- lm(lpsa ~ lcavol + lweight + svi, data=prostate)
lcavol_lweight_svi_lbph_mod <- lm(lpsa ~ lcavol + lweight + svi + lbph, data=prostate)
lcavol_lweight_svi_lbph_age_mod <- lm(lpsa ~ lcavol + lweight + svi + lbph + age, data=prostate)
lcavol_lweight_svi_lbph_age_lcp_mod <- lm(lpsa ~ lcavol + lweight + svi + lbph + age + lcp, data=prostate)
lcavol_lweight_svi_lbph_age_lcp_pgg45_mod <- lm(lpsa ~ lcavol + lweight + svi + lbph + age + lcp + pgg45, data=prostate)
lcavol_lweight_svi_lbph_age_lcp_pgg45_gleason_mod <- lm(lpsa ~ lcavol + lweight + svi + lbph + age + lcp + pgg45 + gleason, data=prostate)

# store the R-squared and RSE values for each model
rsquared_vals <- c(summary(lcavol_mod)$adj.r.squared,
              summary(lcavol_lweight_mod)$adj.r.squared,
              summary(lcavol_lweight_svi_mod)$adj.r.squared,
              summary(lcavol_lweight_svi_lbph_mod)$adj.r.squared,
              summary(lcavol_lweight_svi_lbph_age_mod)$adj.r.squared,
              summary(lcavol_lweight_svi_lbph_age_lcp_mod)$adj.r.squared,
              summary(lcavol_lweight_svi_lbph_age_lcp_pgg45_mod)$adj.r.squared,
              summary(lcavol_lweight_svi_lbph_age_lcp_pgg45_gleason_mod)$adj.r.squared)

rse_vals <- c(summary(lcavol_mod)$sigma,
         summary(lcavol_lweight_mod)$sigma,
         summary(lcavol_lweight_svi_mod)$sigma,
         summary(lcavol_lweight_svi_lbph_mod)$sigma,
         summary(lcavol_lweight_svi_lbph_age_mod)$sigma,
         summary(lcavol_lweight_svi_lbph_age_lcp_mod)$sigma,
         summary(lcavol_lweight_svi_lbph_age_lcp_pgg45_mod)$sigma,
         summary(lcavol_lweight_svi_lbph_age_lcp_pgg45_gleason_mod)$sigma)

# displays the R-squared and RSE for each of these models
lmod_all_outputs <- data.frame(predictors=c('lcavol', 'lcavol_lweight', 'lcavol_lweight_svi', 'lcavol_lweight_svi_lbph', 'lcavol_lweight_svi_lbph_age', 'lcavol_lweight_svi_lbph_age_lcp', 'lcavol_lweight_svi_lbph_age_lcp_pgg45', 'lcavol_lweight_svi_lbph_age_lcp_pgg45_gleason'),
                               R_squared=rsquared_vals, RSE=rse_vals)

print(lmod_all_outputs)
```

<br>

#### Plot the R-squared and RSE trends as predictors are successively added to the model

```{r plots}

# Trend in R-squared as predictors added
plot(lmod_all_outputs$R_squared, xlab = "Number of predictor variables", ylab = "R-squared", main = "R-squared value in each model")

# Trend in RSE as predictors added
plot(lmod_all_outputs$RSE, xlab = "Number of predictor variables", ylab = "Residual Standard Error", main = "RSE value in each model")
```

* As predictor variables are added to the model, R-squared (goodness of fit measure) increases, while the RSE (error or amount of deviation between predicted and actual values) decreases

<br>